---
title: "Continuous Flow Examples"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous Flow Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Isoreader supports several continuous flow IRMS data formats as well as plotting and data conversion functions specifically for continuous flow data. This vignette shows some of the functionality for continuous flow files. For additional information on operations more generally (caching, combining read files, data export, etc.), please consult the [operations vignette](https://kopflab.github.io/isoreader/articles/operations.html).

Note: this vignette is still a work in progress.

```{r}
# load isoreader package
library(isoreader)
```


# Reading files

Reading continuous flow files is as simple as passing one or multiple file or folder paths to the `iso_read_continuous_flow()` function. If folders are provided, any files that have a recognized continuous flow file extensions within those folders will be processed (e.g. all `.dxf`, `.cf` and `.iarc`). Here we read several files that are bundled with the package as examples (and whose paths can be retrieved using the `iso_get_reader_example()` function). Note that some of the files (.cf, .dxf) are individual analysis files whereas others (.iarc) are collections of several files.

```{r}
# all available examples
iso_get_reader_examples() %>% rmarkdown::paged_table()
# read a few of the continuous flow examples
iso_files <- 
  iso_read_continuous_flow(
    iso_get_reader_example("continuous_flow_example.cf"),
    iso_get_reader_example("continuous_flow_example.iarc"),
    iso_get_reader_example("continuous_flow_example.dxf"),
    iso_get_reader_example("linearity_example.dxf")
  )
```

# File summary

The `iso_files` variable now contains a set of isoreader objects, one for each file. Take a look at what information was retrieved from the files using the `iso_get_data_summary()` function.

```{r}
iso_files %>% iso_get_data_summary() %>% rmarkdown::paged_table() 
```

## Problems

In case there was any trouble with reading any of the files, the following functions provide an overview summary as well as details of all errors and warnings, respectively. The examples here contain no errors but if you run into any unexpected file read problems, please file a bug report in the [isoreader issue tracker](https://github.com/kopflab/isoreader/issues).

```{r}
iso_files %>% iso_get_problems_summary() %>% rmarkdown::paged_table()
iso_files %>% iso_get_problems() %>% rmarkdown::paged_table()
```

# File Information

Detailed file information can be aggregated for all isofiles using the `iso_get_file_info()` function which supports the full [select syntax](https://dplyr.tidyverse.org/reference/select.html) of the [dplyr](https://dplyr.tidyverse.org/) package to specify which fields are of interest (by default, all file information is retrieved). The following provides a few examples for how this can be used (the names of the interesting info columns may vary between different file formats):

```{r}
# all file information
iso_files %>% iso_get_file_info(select = c(-file_root)) %>% rmarkdown::paged_table()
# select file information
iso_files %>% 
  iso_get_file_info(
    select = c(
      # select the Analysis column (keep the name)
      Analysis,
      # select the time stamp and rename it to `Date & Time`
      `Date & Time` = file_datetime, 
      # select all columns that start with "Id" and rename them id1, id2, etc.
      id = starts_with("Id"),
      # select all columns that contain "Method"
      matches("Method"))
  ) %>% rmarkdown::paged_table() 
```

## Resistors

Additionally, some IRMS data files contain resistor information that are useful for downstream calculations (see e.g. section on signal conversion later in this vignette):

```{r}
iso_files %>% iso_get_resistors_info() %>% rmarkdown::paged_table()
```

## Reference values

As well as isotopic reference values for the different gases:

```{r}
# reference delta values without ratio values
iso_files %>% iso_get_standards_info() %>% rmarkdown::paged_table()
# reference values with ratios
iso_files %>% iso_get_standards_info(with_ratios = TRUE) %>% rmarkdown::paged_table()
```


# Data & Plots

The raw data read from the IRMS files can be retrieved similarly using the `iso_get_raw_data()` function. Most data aggregation functions also allow for inclusion of file information using the `include_file_info` parameter, which functions identically to the `select` parameter of the `iso_get_file_info` function discussed earlier.

```{r}
# get raw data with default selections (all raw data, no additional file info)
iso_files %>% iso_get_raw_data() %>% head(n=10) %>% rmarkdown::paged_table()
# get specific raw data and add some file information
iso_files %>% 
  iso_get_raw_data(
    # select just time and the m/z 2 and 3 ions
    select = c(time.s, v2.mV, v3.mV),
    # include the Analysis number fron the file info and rename it to 'run'
    include_file_info = c(run = Analysis)
  ) %>% 
  # look at first few records only
  head(n=10) %>% rmarkdown::paged_table()
```


## Chromatograms

Plotting the raw data (i.e. the mass chromatograms) can be done either with the generic `iso_plot_raw_data()` function for a simple plot with default parameters, or directly using the continuous flow specific `iso_plot_continuous_flow()` which can be highly customized. Note that the following plot shows the data from all files in their originally recorded signal intensity units. For examples of how to convert to a common unit, see the section on **Signal conversion** further down.

```{r "all data", fig.width = 10, fig.asp = 2}
iso_files %>% iso_plot_continuous_flow_data()
```

All customization options are described in the function help (`?iso_plot_continuous_flow_data`) and include, for example, plotting only a specific subset of masses (which will omit data from files that don't include this mass):

```{r "select masses", fig.width = 10, fig.asp = 1}
iso_files %>% 
  # plot just masses 2, 44 and 45
  iso_plot_continuous_flow_data(
    data = c("2", "44", "45")
  )
```

## Isotope ratios

Raw data typically only includes ion signals but isotope ratios are often needed to calculate and visualize the data. For this purpose, isoreader provides a dynamic ratio calculation function (`iso_calculate_ratios()`) that accepts any combination of masses, here demonstrated for the standard CO2 ratios $\frac{46}{44}$ and $\frac{45}{44}$. Additionally, the following example demonstrates how the `filter` parameter can be used to exclude plotting artifacts (here e.g. the extreme ratio values seen right after the magnet jump). Notice how the ratios calculated from current intensities are close to 0 (as the real ratios likely are) whereas those calculated from voltages are close to 1 (due to the so chosen, different resistors in the detector circuit). See the section on **Signal conversion** further down on how to scale to uniform intensity units to address this comparison obstacle.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# for between
library(dplyr)
# for expand range
library(ggplot2)
```

```{r "ratios", fig.width = 10, fig.asp = 0.8}
iso_files <-
  iso_files %>% 
  # calculate 46/44 and 45/44 ratios
  iso_calculate_ratios(ratios = c("46/44", "45/44"))

iso_files %>% 
  iso_plot_continuous_flow_data(
    # visualize ratios along with main ion
    data = c("44", "45/44", "46/44"),
    # plot all masses but ratios only 0 to 2 to omit peak jump artifacts
    filter = category == "mass" | between(value, 0, 2)
  ) + 
  expand_limits(y = -0.3)
```

## Time conversion

Most continuous flow data is reported in time units of seconds but this is not always the most useful unit. Isoreader provides easy time scaling functionality that will work for all standard time units (minutes, hours, days, etc.) regardless of which time units individual isofiles were initially recorded in. 

```{r "time conversion", fig.width = 10, fig.asp = 0.5}
iso_files <- 
  iso_files %>% 
  # convert to minutes
  iso_convert_time(to = "minutes") 

iso_files %>% 
  # plot masses 2 and 3
  iso_plot_continuous_flow_data(data = c(2, 3))
```

In this context, another useful customization option of the plotting function is the option to select a specific time window (in any time units, regardless of what is plotted), and the possibility to adjust plot aesthetics such as color, linetype, and paneling:

```{r "time window", fig.width = 10, fig.asp = 0.5}
iso_files %>% 
  # calculate 3/2 ratio on the file
  iso_calculate_ratios("3/2") %>% 
  iso_plot_continuous_flow_data(
    # replot masses 2 and 3 together with the ratio for a specific time window
    data = c("2", "3", "3/2"),
    time_interval = c(10, 13),
    time_interval_units = "minutes",
    # adjust plot aesthetics to panel by masses vs. ratios and color by traces
    panel = category,
    color = data
  )
```

## Signal conversion

Likewise, isoreader can convert between different signal units. This is particularly useful for comparing data files from different mass specs that record signals differentally. The earlier plot of mass 44 illustrated this scenario where some data was recorded in mV and some in nA. Voltages and currents can be scaled within each unit without restriction and with any valid SI prefix  (e.g. from mV to V or nA to pA), however, for conversion from voltage to current or vice-versa, the appropriate resistor values need to be provided or be available from the data files themselves. The following is an example of scaling from voltage to current with the resistor values automatically read from the original data files (see file information section for details). Notice how the mass 44 signal is now in the same units for all files and the ratios are on comparable scales because they are calculated from the same unit intensities:

```{r "signal conversion", fig.width = 10, fig.asp = 0.7}
iso_files %>% 
  # convert all signals to nano ampere
  iso_convert_signals(to = "nA") %>% 
  # re-calculate ratios with the new unuts
  iso_calculate_ratios(ratios = c("46/44", "45/44")) %>% 
  # re-plot same plot as earlier
  iso_plot_continuous_flow_data(
    data = c("44", "45/44", "46/44"),
    filter = 
      category == "mass" |
      (data == "45/44" & between(value, 0.005, 0.03)) | 
      (data == "46/44" & between(value, 0, 0.012))
  )
```

## Plot styling

Since all isoreader plots are standard [ggplot](http://ggplot2.tidyverse.org/) objects, they can be modified with any ggplot commands. For example to modify the themes:

```{r "styling", fig.width = 10, fig.asp = 0.5}
library(ggplot2)
iso_files %>% 
  # replot
  iso_plot_continuous_flow_data(data = c(2,3)) +
  # add vertical dashed line (ggplot functionality)
  geom_vline(xintercept = 15, linetype = 2) +
  # modify plot styling (ggplot functionality)
  theme(text = element_text(size = 20),
        legend.position = "none")
```

# Data Processing

The isoreader package is intended to make raw stable isotope data easily accessible. However, as with most analytical data, there is significant downstream processing required to turn these raw intensity chromatograms into peak-specific, properly referenced isotopic measurements. This and similar functionality will be part of the [isoprocessor package](https://github.com/KopfLab/isoprocessor) which takes isotopic data through the various corrections in a transparent, efficient and reproducible manner. 

That said, most vendor software also performs some of these calculations and it can be useful to be able to compare new data reduction procecures against those implemented in the vendor software. For this purpose, isoreader retrieves vendor computed data tables whenver possible, as illustrated below.

## Vendor Data Table

As with most data retrieval funtions, the `iso_get_vendor_data_table()` function also allows specific column selection (by default, all columns are selected) and easy addition of file information via the `include_file_info` parameter (by default, none is included).

```{r}
# entire vendor data table
iso_files %>% iso_get_vendor_data_table() %>% rmarkdown::paged_table()
# get specific parts and add some file information
iso_files %>% 
  iso_get_vendor_data_table(
    # select peak number, ret. time, overall intensity and all H delta columns
    select = c(Nr., Rt, area = `rIntensity All`, matches("^d \\d+H")),
    # include the Analysis number fron the file info and rename it to 'run'
    include_file_info = c(run = Analysis)
  ) %>% rmarkdown::paged_table() 
# the data table also provides units if included in the original data file
# caution however: it may require changes to the `select` parameter
iso_files %>% 
  iso_get_vendor_data_table(
    with_units = TRUE,
    # select peak number, ret. time, overall intensity and all H delta columns
    select = c(Nr., matches("^Rt"), matches("rIntensity All"), matches("^d \\d+H")),
    # include the Analysis number fron the file info and rename it to 'run'
    include_file_info = c(run = Analysis)
  ) %>% rmarkdown::paged_table() 
```

## For expert users: retrieving all data 

For users familiar with the nested data frames from the [tidyverse](https://www.tidyverse.org/) (particularly [tidyr](https://tidyr.tidyverse.org/)'s `nest` and `unnest`), there is an easy way to retrieve all data from the iso file objects in a single nested data frame. Use the `include_file_info`, `include_raw_data`, and `include_vendor_data_table` parameters to specify which columns to include. By default, everything is included:

```{r}
all_data <- iso_files %>% iso_get_data()
all_data %>% rmarkdown::paged_table()
```

# Saving collections

Saving entire collections of isofiles for retrieval at a later point is easily done using the `iso_save` function which stores collections or individual isoreader file objects in the efficient R data storage format `.rds` (if not specified, the extension `.cf.rds` will be automatically appended). These saved collections can be convientiently read back using the same `iso_read_continuous_flow` command used for raw data files.

```{r}
# export to R data archive
iso_files %>% iso_save("iso_files_export.cf.rds")

# read back the exported R data archive
iso_files <- iso_read_continuous_flow("iso_files_export.cf.rds")
iso_files %>% iso_get_data_summary() %>% rmarkdown::paged_table()
```


# Data Export

At the moment, isoreader supports export of all data to R Data archives, Excel, and the [Feather file format](https://blog.rstudio.com/2016/03/29/feather/) (a Python/R cross-over format). Note that both export methods have similar syntax and append the appropriate file extension for each type of export file (`.cf.xlsx` and `.cf.feather`, respectively).

```{r}
# export to excel
iso_files %>% iso_export_to_excel("iso_files_export")

# data sheets available in the exported data file:
readxl::excel_sheets("iso_files_export.cf.xlsx")
```

```{r}
# export to feather
iso_files %>% iso_export_to_feather("iso_files_export")

# exported feather files
list.files(pattern = ".cf.feather")
```


