---
title: "Operations Examples"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Coming soon...

Isoreader provides a number of general purpose operations that work on all supported IRMS data formats such as caching of read files and data exports. This will demonstrate some of these general operations.

## Supported file types

```{r}

```


## Caching

```{r}

```


## Messages and default parameters

```{r}

```


## Combining / subsetting isofiles

All isoreader objects are lists that can be combined or subset to work with only specific files or create a larger collection. 

```{r}

```

## Filtering isofiles by file information

Additionally, the `iso_filter_files` function allows for subsetting a collection of isofiles based on any file information.

```{r}

```

## Removing erroneous files

Removing files with problems: `iso_omit_files_with_problems`

```{r}

```

## Re-reading files

```{r}

```




