---
title: "Quick Start Guide"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options:
  chunk_output_type: console
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Isoreader supports various dual inlet, continuous flow, and scan file formats. This vignette shows how to get started reading these raw IRMS data files and exporting the information to Excel. For more details on isoreader functionality and each file type, please read the **Full Examples** vignettes. For more information on downstream processing with isoverse, check out the [isoprocessor](https://isoprocessor.isoverse.org) package.

```{r, message=FALSE}
# load isoreader package
library(isoreader)
```

```{r, include=FALSE, message=FALSE}
iso_turn_reader_caching_off()
```


# Data files

For demonstration purposes, this vignette simply reads all supported dual inlet, continuous flow, and scan files that are bundled with the isoreader package. For a more detailed introduction to each file type, check out the specific vignettes for each:

 - [Continuous Flow](https://isoreader.isoverse.org/articles/continuous_flow.html)
 - [Dual Inlet](https://isoreader.isoverse.org/articles/dual_inlet.html)
 - [Scans](https://isoreader.isoverse.org/articles/scan.html)


```{r}
# all available examples
iso_get_reader_examples() %>% knitr::kable()
```

# Dual Inlet Files

```{r}
# read all available examples
di_files <- iso_read_dual_inlet(iso_get_reader_examples_folder())
```

```{r}
# save as r data storage (read back in with iso_read_dual_inlet)
iso_save(di_files, filepath = "di_save")
```

```{r, eval = FALSE}
# export to excel
iso_export_to_excel(di_files, filepath = "di_export")
```

# Continuous Flow Files

```{r}
# read all available examples
cf_files <- iso_read_continuous_flow(iso_get_reader_examples_folder())
```

```{r}
# save as r data storage (read back in with iso_read_continuous_flow)
iso_save(cf_files, filepath = "cf_save")
```

```{r, eval = FALSE}
# export to excel
iso_export_to_excel(cf_files, filepath = "cf_export")
```

# Scan Files

```{r}
# read all available examples
scan_files <- iso_read_scan(iso_get_reader_examples_folder())
```

```{r}
# save as r data storage (read back in with iso_read_scan)
iso_save(scan_files, filepath = "scan_save")
```

```{r, eval = FALSE}
# export to excel
iso_export_to_excel(scan_files, filepath = "scan_export")
```

